---
title: "Pay Per Ride or Buy a Pass? A SEPTA Break-Even Guide"
author: "Yining Han"
date: 2026-09-21
categories:
  - web scraping
  - transportation
  - personal finance
description: "I scrape SEPTA's fare page to find the exact number of Bus and Metro trips that makes each pass worth buying."
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup-and-scrape

library(rvest)
library(httr2)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(knitr)

fare_url <- "https://wwww.septa.org/fares/"
snapshot_path <- "data/septa_fares_snapshot.csv"

scrape_septa_fares <- function(url) {
  # This project makes only one page request. The short pause is an
  # additional courtesy if the function is sourced in a larger workflow.
  Sys.sleep(0.5)

  page <- request(url) |>
    req_user_agent("Academic-Web-Scraping-Project/1.0") |>
    req_timeout(30) |>
    req_perform() |>
    resp_body_raw() |>
    read_html()

  heading_text <- page |>
    html_elements("h2") |>
    html_text2()

  pay_per_ride_text <- heading_text[
    str_detect(heading_text, "^\\$[0-9]+(?:\\.[0-9]+)?/Trip$")
  ]

  list_text <- page |>
    html_elements("li") |>
    html_text2()

  pass_text <- list_text |>
    str_subset(
      "One Day Convenience Pass|Three Day Convenience Pass|Weekly TransPass\\+|Monthly TransPass\\+"
    ) |>
    unique()

  if (length(pay_per_ride_text) == 0 || length(pass_text) != 4) {
    stop("The SEPTA page structure has changed; update the selectors.")
  }

  get_price <- function(x) {
    as.numeric(str_remove(str_extract(x, "\\$[0-9]+(?:\\.[0-9]+)?"), "\\$"))
  }

  pass_data <- tibble(
    product = case_when(
      str_detect(pass_text, "One Day") ~ "One Day Convenience Pass",
      str_detect(pass_text, "Three Day") ~ "Three Day Convenience Pass",
      str_detect(pass_text, "Weekly") ~ "Weekly TransPass+",
      str_detect(pass_text, "Monthly") ~ "Monthly TransPass+"
    ),
    price = get_price(pass_text),
    validity = case_when(
      str_detect(pass_text, "One Day") ~ "One day",
      str_detect(pass_text, "Three Day") ~ "Three-day period",
      str_detect(pass_text, "Weekly") ~ "One week",
      str_detect(pass_text, "Monthly") ~ "One month"
    )
  )

  bind_rows(
    tibble(
      product = "Pay per ride",
      price = get_price(pay_per_ride_text[[1]]),
      validity = "One paid Bus or Metro trip"
    ),
    pass_data
  ) |>
    mutate(
      scraped_on = as.character(Sys.Date()),
      source_url = url
    )
}

# Keep a dated snapshot so the post remains reproducible if the live page is
# temporarily unavailable. A successful scrape refreshes that snapshot.
fares_live <- tryCatch(
  scrape_septa_fares(fare_url),
  error = function(e) NULL
)

if (is.null(fares_live)) {
  fares <- read.csv(snapshot_path, stringsAsFactors = FALSE)
  data_status <- "the repository snapshot"
} else {
  fares <- fares_live
  write.csv(fares, snapshot_path, row.names = FALSE)
  data_status <- "a live scrape"
}

data_date <- max(as.Date(fares$scraped_on))

ride_fare <- fares |>
  filter(product == "Pay per ride") |>
  pull(price)

break_even <- fares |>
  filter(product != "Pay per ride") |>
  mutate(
    # Work in integer cents so an exact tie is not distorted by
    # floating-point rounding.
    trips_to_save = round(price * 100) %/% round(ride_fare * 100) + 1,
    pay_per_ride_cost = trips_to_save * ride_fare,
    savings_at_threshold = pay_per_ride_cost - price
  )

get_threshold <- function(pass_name) {
  break_even |>
    filter(product == pass_name) |>
    pull(trips_to_save)
}

one_day_threshold <- get_threshold("One Day Convenience Pass")
three_day_threshold <- get_threshold("Three Day Convenience Pass")
weekly_threshold <- get_threshold("Weekly TransPass+")
monthly_threshold <- get_threshold("Monthly TransPass+")
monthly_tie <- break_even |>
  filter(product == "Monthly TransPass+") |>
  pull(price) / ride_fare
```

SEPTA offers a simple choice that is surprisingly easy to get wrong: pay for each Bus or Metro trip, or buy a pass in advance. A pass feels economical because it offers many trips for a fixed price, but it can cost more when a rider does not use it enough. I therefore ask a practical question: **how many paid trips make each SEPTA pass cheaper than paying per ride?** The answer can guide commuters, students, and visitors before they load a SEPTA Key card.

## Data and responsible scraping

I collected the prices from SEPTA's public [Fare Information page](https://www.septa.org/fares/). The data used here are dated `r data_date`. SEPTA serves the readable static HTML at its own `wwww.septa.org` subdomain, while the reader-facing link uses the usual `www` address. The page requires no account, paywall, CAPTCHA, or other access control. When I accessed it, its `robots.txt` permitted the extension-free fare page. My script identifies itself with a descriptive user agent, pauses briefly, and makes only one request. It then uses `rvest` to extract the pay-per-ride heading and the four Bus and Metro pass entries. The analysis currently uses `r data_status`; the dated CSV snapshot in the repository is the fallback if SEPTA is temporarily unavailable.

The scraper selects the relevant `<h2>` and `<li>` elements, cleans dollar signs and descriptive text, and converts each price to a number. I also included checks that stop the live scrape if SEPTA changes the page structure, rather than silently returning incorrect data.

## Turning prices into a decision rule

For a pass costing $P$ and a pay-per-ride fare of $f$, the first trip count at which the pass is *strictly* cheaper is

$$
\left\lfloor \frac{P}{f} \right\rfloor + 1.
$$

This counts **paid trips**, not every vehicle boarded. SEPTA currently allows up to two free transfers within two hours when riders use the same Key card or contactless card, so a bus-to-Metro journey with a free transfer still counts as one paid trip.

```{r}
#| label: tbl-break-even
#| tbl-cap: "The first number of paid trips at which each pass costs less than pay per ride."

break_even |>
  transmute(
    Pass = product,
    Price = dollar(price),
    `First trip count where pass is cheaper` = trips_to_save,
    `Savings at that count` = dollar(savings_at_threshold)
  ) |>
  kable(align = c("l", "r", "r", "r"))
```

The short-period passes have intuitive thresholds. The one-day pass pays off on paid trip `r one_day_threshold`, while the three-day pass pays off on trip `r three_day_threshold`. The weekly result is more surprising: a standard five-day round-trip commute is 10 paid trips and costs `r dollar(10 * ride_fare)`, which is still `r dollar(break_even$price[break_even$product == "Weekly TransPass+"] - 10 * ride_fare)` less than the weekly pass. The pass first wins at `r weekly_threshold` trips, so one additional paid trip during the week changes the decision.

The monthly pass has an exact tie at `r monthly_tie` trips. A rider making 20 round-trip commute days pays `r dollar(40 * ride_fare)` either way; the pass begins saving money on trip `r monthly_threshold`.

```{r}
#| label: fig-monthly-break-even
#| fig-cap: "The monthly pass and pay-per-ride cost lines cross at the break-even point."
#| fig-width: 7
#| fig-height: 4.5

monthly_price <- break_even |>
  filter(product == "Monthly TransPass+") |>
  pull(price)

monthly_costs <- tibble(
  paid_trips = 0:60,
  pay_per_ride = paid_trips * ride_fare,
  monthly_pass = monthly_price
)

monthly_plot <- ggplot(monthly_costs, aes(x = paid_trips)) +
  geom_line(
    aes(y = pay_per_ride, color = "Pay per ride"),
    linewidth = 1.1
  ) +
  geom_line(
    aes(y = monthly_pass, color = "Monthly pass"),
    linewidth = 1.1
  ) +
  geom_vline(
    xintercept = monthly_price / ride_fare,
    color = "#666666",
    linetype = "dotted"
  ) +
  annotate(
    "point",
    x = monthly_price / ride_fare,
    y = monthly_price,
    size = 2.8
  ) +
  annotate(
    "text",
    x = monthly_price / ride_fare + 1.5,
    y = monthly_price - 9,
    label = paste0("Equal at ", monthly_tie, " trips"),
    hjust = 0
  ) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_manual(
    values = c("Pay per ride" = "#1f77b4", "Monthly pass" = "#d95f02")
  ) +
  labs(
    x = "Paid Bus or Metro trips in one month",
    y = "Total fare cost",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

dir.create("figures", showWarnings = FALSE)
ggsave(
  "figures/monthly_break_even.png",
  monthly_plot,
  width = 7,
  height = 4.5,
  dpi = 160
)

monthly_plot
```

## Recommendation

The actionable rule is simple: **pay per ride when travel is irregular; buy a pass only when the expected paid-trip count reaches its threshold.** For Bus and Metro travel alone, a five-day round-trip commuter should pay per ride for a 10-trip week, but choose the weekly pass at `r weekly_threshold` or more trips. A monthly commuter should buy the pass at `r monthly_threshold` or more trips, while a visitor making at least `r one_day_threshold` paid trips in one day should choose the one-day pass.

This comparison is deliberately conservative. Passes can include Regional Rail privileges, and some riders value cost certainty even at a small premium. Discounts and special programs may also change the calculation. Still, the central takeaway remains: a fixed-price pass is not automatically a bargain. Count paid trips first, remember that free transfers do not add fares, and rerun the scraper whenever SEPTA updates its prices.

## Reproducibility

All reported values, the table, and the figure are generated from the scraped data. Package and platform details are recorded below.

```{r}
#| label: session-info
#| code-fold: true

sessionInfo()
```
